---
title: "Web Traffic Model Example"
author: "Harry He"
date: "2026-07-26"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Web Traffic Model Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Purpose of vignette

This vignette illustrates `capnet` in a domain other than finance: forecasting
daily aggregate Wikipedia traffic from a panel of individual, heavy-tailed
article-level pageview series. Article popularity on the web is famously
Zipf/heavy-tailed -- a small number of articles can go "viral" and receive
orders of magnitude more traffic than usual on a given day. A linear (here,
Poisson) forecasting model that regresses tomorrow's site-wide traffic on
today's per-article pageviews is exactly the setting `capnet` targets: without
a contribution cap, a single freak spike in one article's traffic can swamp
the forecast, even though the underlying coefficient is not unusually large
under a standard elastic-net fit.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Core package (required)
library(capnet)

# Optional packages for vignette
needed <- c(
  "glmnet", "lbfgs", "ggplot2",
  "dplyr", "tidyr", "purrr", "lubridate", "zoo"
)

missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing) > 0) {
  knitr::opts_chunk$set(eval = FALSE)
  message(
    "Optional packages not installed: ",
    paste(missing, collapse = ", "),
    ". Code chunks will not be evaluated."
  )
} else {
  library(glmnet)
  library(lbfgs)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(lubridate)
  library(zoo)
}

if (length(missing) == 0) {
  source("helpers.R", local = knitr::knit_global())
}
```

### Data

The dataset (`inst/extdata/wiki_traffic.csv`) contains daily English Wikipedia
pageviews (2022-01-01 to 2024-12-31) for 30 widely-read articles (major tech
companies, AI/crypto topics, and prominent public figures), pulled from the
public Wikimedia Pageviews REST API. The response is *next-day* total
pageviews summed across all 30 articles; the covariates are *today's*
per-article pageviews (in thousands of views, for numerical scaling). The
`total` column is dropped from the covariate set since it is an exact linear
combination of the 30 article columns.

```{r data, result="asis"}
wiki <- read.csv(system.file("extdata", "wiki_traffic.csv", package = "capnet"))
wiki$date <- as.Date(wiki$date)

articles <- setdiff(names(wiki), c("date", "total"))

wiki <- wiki |>
  arrange(date) |>
  mutate(target = dplyr::lead(total, 1)) |>
  filter(!is.na(target)) |>
  mutate(across(all_of(articles), ~ . / 1000)) |>   # thousands of views
  select(date, target, all_of(articles))

print(calc_moments(wiki[, articles]))
```

The summary statistics show pronounced positive excess kurtosis for several
articles (e.g. `sam_altman`, `elon_musk`, `nvidia`), confirming that a handful
of articles experience occasional traffic spikes far outside their typical
range -- precisely the covariate behavior contribution caps are designed to
control.

### Backtest

We forecast next-day total pageviews with a Poisson log-link model, using a
rolling one-year training window refit monthly (steps of 5 days within each
month, to keep runtime moderate) and evaluated out-of-sample from 2023-01-01
through 2024-12-31. Refit months are independent of one another, so
`rolling_enet()` dispatches them across CPU cores via `parallel = TRUE`.
Because Poisson deviance scales with the (very large) raw
pageview counts, we set `alpha = 0` and forgo elastic-net selection entirely
(`lambda = 0`) so that the comparison isolates the effect of the contribution
cap alone, exactly mirroring the role `gamma` plays in the finance example.

Contribution ceilings are computed the same way as in the finance vignette:
as a multiple of each feature's univariate regression coefficient with the
response, here using a univariate Poisson GLM slope (the natural analogue of
the OLS slope used for the Gaussian model) rather than the raw correlation.

```{r experiment}
wiki_L <- function(train_use, target_col, feature_cols, y_train) {
  sapply(feature_cols, function(v) {
    uni <- suppressWarnings(
      stats::glm(train_use[[target_col]] ~ train_use[[v]], family = "poisson")
    )
    3 * abs(stats::coef(uni)[2])
  })
}

results <- rolling_enet(
  wiki,
  date_col = "date",
  target_col = "target",
  alpha = 0,
  window_years = 1,
  method = c("capnet_walk", "capnet", "postcap"),
  capnet_gamma = c(0),
  walk_gamma = c(10, 100, 1000, 10000, 100000),
  min_obs = 200,
  eval_start = "2023-01-01",
  family = "poisson",
  cap_formula = wiki_L,
  lambda_override = 0,
  walk = 5,
  parallel = TRUE
)

combos <- results$predictions %>%
  dplyr::distinct(method, gamma) %>%
  dplyr::arrange(method, gamma)

predictions <- list()
contributions <- list()
methods <- c()
gamma_vals <- c()

for (i in seq_len(nrow(combos))) {
  m <- combos$method[i]
  gamma_j <- combos$gamma[i]

  predictions[[i]] <- results$predictions %>%
    dplyr::filter(method == m, gamma == gamma_j)
  contributions[[i]] <- results$contributions %>%
    dplyr::filter(method == m, gamma == gamma_j)
  methods[i] <- m
  gamma_vals[i] <- gamma_j
}
```

```{r summary, echo = FALSE}
summary_table <- lapply(seq_along(methods), function(i) {
  tibble::tibble(
    method = methods[i],
    gamma = ifelse(gamma_vals[i] == 0, NA_real_, gamma_vals[i]),
    oos_r2 = calc_oos_r2(predictions[[i]], date_col = "date")
  )
}) |>
  dplyr::bind_rows()

print(summary_table)
```

The uncapped Poisson model is catastrophically unstable out-of-sample: with
30 correlated daily-traffic covariates and no regularization, it occasionally
produces an enormous predicted count when a covariate combination lands far
outside the training distribution, driving its out-of-sample pseudo-R^2 deeply
negative. Post-fit capping clips the worst contributions after the fact but
does not let the coefficients adjust, so it remains unstable (pseudo-R^2 still
negative). The walk-forward `capnet` model, which incorporates the
contribution cap directly into estimation, is dramatically more stable across
the entire range of `gamma` values tested and is the only approach with a
positive out-of-sample pseudo-R^2.

```{r fig.height = 6.5, fig.width = 10, echo = FALSE}
best_gamma <- summary_table |>
  dplyr::filter(method == "Walk-forward Capnet") |>
  dplyr::slice_max(oos_r2, n = 1) |>
  dplyr::pull(gamma)

sel <- which(
  (methods == "Post-Fit Cap") |
  (methods == "Uncapped") |
  (methods == "Walk-forward Capnet" & gamma_vals == best_gamma)
)

rolling_list <- lapply(sel, function(i) {
  df <- rolling_oos_r2(predictions[[i]], window = 180, date_col = "date") |>
    tidyr::drop_na()
  df$spec <- paste0(
    methods[i],
    ifelse(gamma_vals[i] == 0, "", paste0(" (gamma=", gamma_vals[i], ")"))
  )
  df
})

plot_df <- dplyr::bind_rows(rolling_list) %>%
  dplyr::mutate(date = as.Date(date))

walk_label <- paste0("Walk-forward Capnet (gamma=", best_gamma, ")")

ggplot(plot_df, aes(x = date, y = r2_roll, color = spec, group = spec)) +
  geom_line(linewidth = 0.8, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = setNames(
    c("steelblue", "firebrick", "darkolivegreen4"),
    c("Uncapped", "Post-Fit Cap", walk_label)
  )) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y-%m") +
  coord_cartesian(ylim = c(-3, 1)) +
  labs(
    x = "Date", y = expression(R^2), color = "Model",
    caption = sprintf(
      "Y-axis truncated for readability; Uncapped's rolling R^2 reaches a minimum of %.0f.",
      min(plot_df$r2_roll[plot_df$spec == "Uncapped"])
    )
  ) +
  theme_bw(base_size = 16) +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 0)
  )
```

```{r fig.height = 6.5, fig.width = 10, echo = FALSE}
ref <- contributions[[sel[1]]]
feat_start <- which(names(ref) == "lambda") + 1
max_col_idx <- max.col(abs(ref[, feat_start:ncol(ref)]), ties.method = "first")

max_contributions <- lapply(sel, function(i) {
  df <- contributions[[i]]
  mat <- as.matrix(df)
  max_col <- as.numeric(mat[cbind(seq_len(nrow(mat)), max_col_idx + feat_start - 1)])
  df <- df[, 1:(feat_start - 1)]
  df$max_col <- max_col
  df$spec <- paste0(
    methods[i],
    ifelse(gamma_vals[i] == 0, "", paste0(" (gamma=", gamma_vals[i], ")"))
  )
  df
})

plot_df <- dplyr::bind_rows(max_contributions) %>%
  dplyr::mutate(date = as.Date(date))

ggplot(plot_df, aes(x = date, y = max_col, color = spec, group = spec)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = setNames(
    c("steelblue", "firebrick", "darkolivegreen4"),
    c("Uncapped", "Post-Fit Cap", walk_label)
  )) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y-%m") +
  labs(x = "Date", y = "Maximum |contribution| (log-mean scale)", color = "Model") +
  theme_bw(base_size = 16) +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 0)
  )
```

The maximum-contribution series shows the mechanism behind the stability
difference: the uncapped model occasionally lets a single article's lagged
pageviews swing the log-mean forecast by several units (a multiplicative
factor of $e^{|\cdot|}$ on the predicted count -- catastrophic for a Poisson
mean), while the walk-forward `capnet` model keeps this quantity close to the
cap throughout, redistributing predictive weight across the panel instead of
concentrating it in one article.

### Contribution redistribution snapshot

```{r, include = FALSE}
make_feature_table <- function(df_list, target_date, df_names = NULL,
                                feature_cols) {
  if (is.null(df_names)) {
    df_names <- paste0("df", seq_along(df_list))
  }

  long_list <- lapply(seq_along(df_list), function(i) {
    df_list[[i]] %>%
      dplyr::filter(date == target_date) %>%
      dplyr::select(dplyr::all_of(feature_cols)) %>%
      tidyr::pivot_longer(
        cols = dplyr::everything(),
        names_to = "feature",
        values_to = df_names[i]
      )
  })

  out <- Reduce(function(x, y) dplyr::left_join(x, y, by = "feature"), long_list)
  as.data.frame(out)
}
```

`2023-03-14` is the single day responsible for the uncapped model's worst
out-of-sample blowup: a spike in `youtube`'s lagged pageviews (relative to
that fit's training distribution) pushes its uncapped log-mean contribution
far above any other feature, producing an implausible next-day forecast.

```{r contrib_table, result="asis"}
tab <- make_feature_table(
  df_list = list(contributions[[which(methods == "Uncapped")[1]]],
                 contributions[[which(methods == "Post-Fit Cap")[1]]],
                 contributions[[sel[length(sel)]]]),
  target_date = as.Date("2023-03-14"),
  df_names = c("Uncapped", "Post-Fit Cap", walk_label),
  feature_cols = articles
)

print(tab)
```

As in the finance example, the post-fit approach truncates the offending
contribution mechanically without adjusting any other coefficient, while the
walk-forward `capnet` model redistributes weight across multiple articles,
reflecting a genuinely re-estimated (rather than post-hoc clipped) fit.
