---
title: "Capping Variable Contributions"
output: pdf_document
date: "2025-06-17"
---

### Purpose of Vignette
This vignette demonstrates how \texttt{capnet} works in practice and how it affects model performance in different contexts. Traditional regularization methods like elastic net penalize coefficients, but in many real-world datasets, variables can have unstable variance or sporadic large magnitudes, meaning that a few predictors can dominate the fitted model. By explicitly capping each variable's maximum contribution $|x_j\beta_j|$, \texttt{capnet} provides a way to directly control how much influence each predictor can exert. This vignette compares standard elastic net, the standard \texttt{capnet} model, and a walk-forward \texttt{capnet} model designed for time-series data and greater stability.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, include=FALSE}
library(glmnet)
library(lbfgs)
library(xts)
library(knitr)
source("../R/capnet.R")
source("../R/cv_capnet.R")
source("../R/walk_capnet.R")
```

```{r, include=FALSE}
simulate_simple_proxy <- function(
  n = 1200, p = 8,
  sparsity = 0.5,          # zeros in beta (on Z -> Y)
  z_sd = 1,                # scale of Z
  eps_sd = 1,              # noise in Y
  unstable_idx = NULL,     # which X columns have unstable variance
  unstable_frac = 0.25,    # used only if unstable_idx is NULL
  a_sd = 0.3,              # multiplicative scaling variation a_j
  sigma_low = 1.0,         # base noise sd for X
  sigma_high = 8.0,        # high noise sd for unstable features
  high_prob = 0.10,        # P(sigma_high) per observation for unstable features
  seed = 2025
) {
  set.seed(seed)

  # 1) Latent drivers Z
  Z <- matrix(rnorm(n * p, sd = z_sd), n, p)

  # 2) True beta and outcome Y
  beta <- rnorm(p)
  if (sparsity > 0) beta[sample(1:p, floor(sparsity * p))] <- 0
  y <- drop(Z %*% beta + rnorm(n, sd = eps_sd))

  # 3) Linear link from Z to X with per-feature scale a_j
  a <- exp(rnorm(p, mean = 0, sd = a_sd))  # mild variation
  mean_part <- sweep(Z, 2, a, `*`)

  # 4) Decide which columns in X have unstable variance
  if (is.null(unstable_idx)) {
    k <- max(1, round(unstable_frac * p))
    unstable_idx <- sort(sample(seq_len(p), k, replace = FALSE))
  }
  stable_idx <- setdiff(seq_len(p), unstable_idx)

  # 5) Build noise matrix Eta
  Eta <- matrix(rnorm(n * p, sd = sigma_low), n, p)  # start with stable noise
  if (length(unstable_idx) > 0) {
    # For unstable cols, each row flips to high variance with prob = high_prob
    flips <- matrix(rbinom(n * length(unstable_idx), 1, high_prob), n, length(unstable_idx))
    sd_mat <- sigma_low + flips * (sigma_high - sigma_low)
    Eta[, unstable_idx] <- matrix(rnorm(n * length(unstable_idx)), n, length(unstable_idx)) * sd_mat
  }

  # 6) Observed proxies: X = a*Z + Eta
  X <- mean_part + Eta

  list(X = X, y = y, Z = Z, beta = beta, a = a,
       unstable_idx = unstable_idx, stable_idx = stable_idx)
}
```

```{r include=FALSE}
prepare_data <- function(data_source = c("quickstart","proxy")) {
  data_source <- match.arg(data_source)
  if (data_source == "quickstart") {
    data(QuickStartExample, package = "glmnet")
    X <- QuickStartExample$x[, 1:5]
    y <- QuickStartExample$y
  } else {
    sim <- simulate_simple_proxy(
      n = 1200, p = 8, unstable_frac = 0.10,
      sigma_low = 1.0, sigma_high = 8.0, high_prob = 0.10, seed = 2025)
    X <- sim$X
    y <- sim$y
  }
  y <- scale(y)
  set.seed(123)
  shuffle <- sample(seq_len(nrow(X)))
  X <- X[shuffle, ]
  y <- y[shuffle, , drop = FALSE]
  
  n <- nrow(X)
  split <- floor(n * 0.8)
  train_X <- X[1:split, ]
  train_y <- y[1:split, , drop = FALSE]
  test_X <- X[(split + 1):n, ]
  test_y <- y[(split + 1):n, , drop = FALSE]
  
  mu <- 100
  if (data_source == "quickstart") {
    L <- runif(ncol(X), 0.8, 1.2)
  } else {
    kappa <- 0.1
    L <- kappa * apply(abs(train_X), 2, function(v) stats::quantile(v, 0.99))
  }
  list(train_X = train_X, train_y = train_y, test_X = test_X, test_y = test_y, L = L, mu = mu)
}
```

```{r include = FALSE}
fit_models <- function(train_X, train_y, test_X, L, mu) {
  alphas <- c(0, 0.2, 0.4, 0.6, 0.8, 1)
  best_alpha <- 0
  best_lambda <- 0
  best_loss <- Inf
  
  for (alpha in alphas) {
    enet.cv <- cv.glmnet(train_X, train_y, alpha=alpha)
    lambda_enet <- enet.cv$lambda.min
    enet.x <- glmnet(train_X, train_y, alpha=alpha, lambda=lambda_enet)
    loss <- sum(predict(enet.x, train_X) - train_y) ^ 2
    if (loss < best_loss) {
      best_alpha <- alpha
      best_lambda <- lambda_enet
      best_loss <- loss
    }
  }
  best_enet <- glmnet(train_X, train_y, alpha=best_alpha, lambda=best_lambda)
  enet_beta0 <- coef(best_enet)[1]
  enet_beta <- coef(best_enet)[-1]
  original_pred_mean <- sum(enet_beta * colMeans(train_X))
  for (i in seq_along(enet_beta)) {
    if (abs(enet_beta[i] * max(train_X[,i])) > L[i]) {
      enet_beta[i] <- (L[i] / max(train_X[, i])) * sign(enet_beta[i])
    }
  }
  new_pred_mean <- sum(enet_beta * colMeans(train_X))
  enet_beta0 <- enet_beta0 + (original_pred_mean - new_pred_mean)
  
  cap1 <- capnet(train_X, train_y, best_lambda, best_alpha, mu, L, standardize=FALSE)
  cap2 <- capnet.walk(train_X, train_y, best_lambda, best_alpha, mu, L, test_X, standardize=FALSE)
  
  list(
    enet = list(beta0 = enet_beta0, beta = enet_beta),
    cap1 = list(beta0 = cap1$a0, beta = cap1$beta),
    cap2 = cap2
  )
}

mse_from_beta <- function(X, y, beta0, beta) {
  pred <- as.numeric(beta0 + X %*% beta)
  mean((pred - as.numeric(y)) ^ 2)
}

mse_from_pred <- function(pred, y) {
  mean((as.numeric(pred) - as.numeric(y)) ^ 2)
}
```

### Standard Well-Behaved Data
The first example uses quick start example data from \texttt{glmnet}. Theoretically, we expect all models to perform similarly as all features are well-behaving and should rarely trigger large contribution penalties in the \texttt{capnet} model. 


```{r}
ds_qs <- prepare_data("quickstart")
fits_qs <- fit_models(ds_qs$train_X, ds_qs$train_y, ds_qs$test_X, ds_qs$L, ds_qs$mu)

qs_enet_is <- mse_from_beta(ds_qs$train_X, ds_qs$train_y, fits_qs$enet$beta0, fits_qs$enet$beta)
qs_cap1_is <- mse_from_beta(ds_qs$train_X, ds_qs$train_y, fits_qs$cap1$beta0, fits_qs$cap1$beta)

qs_enet_oos <- mse_from_beta(ds_qs$test_X, ds_qs$test_y, fits_qs$enet$beta0, fits_qs$enet$beta)
qs_cap1_oos <- mse_from_beta(ds_qs$test_X, ds_qs$test_y, fits_qs$cap1$beta0, fits_qs$cap1$beta)
qs_cap2_oos <- mse_from_pred(fits_qs$cap2$predictions, ds_qs$test_y)

report_qs <- data.frame(
  Model = c("Elan", "Optimized Cap", "Walk Cap"),
  IS = c(qs_enet_is, qs_cap1_is, NA),
  OOS = c(qs_enet_oos, qs_cap1_oos, qs_cap2_oos)
)
kable(report_qs, digits = 6)
```

The in sample (IS) and out of sample (OOS) mean squared errors (MSE) show that the elastic net model and the standard \texttt{capnet} model have similar IS and OOS performances. The walk forward model achieved better OOS performance, which can be attributed to it being more adapted to each out of sample observation. This comes at higher time complexity and the absence of a single trained model that best capture the relationship between the feature and target variables.

### Synthetic Data (Unstable Variance Features)
To illustrate the strength of the \texttt{capnet} model, I construct a synthetic dataset based on the latent data-generating process $Z\rightarrow Y$. In this setup, the true predictors $Z$ drive the outcome, but only a noisy proxy $X$ is observed. Importantly, some variables in $X$ exhibit unstable and high variance, occasionally taking on extreme values. Under these conditions, both the Optimized Cap and Walk Cap models outperform the standard elastic net, with the walk-forward implementation achieving the greatest improvement in generalization. These results demonstrate how \texttt{capnet}’s contribution-based constraints effectively limit the influence of high-variance features that might otherwise distort model predictions.

```{r}
ds_px <- prepare_data("proxy")
fits_px <- fit_models(ds_px$train_X, ds_px$train_y, ds_px$test_X, ds_px$L, ds_px$mu)

px_enet_is <- mse_from_beta(ds_px$train_X, ds_px$train_y, fits_px$enet$beta0, fits_px$enet$beta)
px_cap1_is <- mse_from_beta(ds_px$train_X, ds_px$train_y, fits_px$cap1$beta0, fits_px$cap1$beta)

px_enet_oos <- mse_from_beta(ds_px$test_X, ds_px$test_y, fits_px$enet$beta0, fits_px$enet$beta)
px_cap1_oos <- mse_from_beta(ds_px$test_X, ds_px$test_y, fits_px$cap1$beta0, fits_px$cap1$beta)
px_cap2_oos <- mse_from_pred(fits_px$cap2$predictions, ds_px$test_y)

report_px <- data.frame(
  Model = c("Elan", "Optimized Cap", "Walk Cap"),
  IS = c(px_enet_is, px_cap1_is, NA),
  OOS = c(px_enet_oos, px_cap1_oos, px_cap2_oos)
)
kable(report_px, digits = 6)
```








